---
title: "Technical Appendix"
output: 
  html_document:
    toc: true
    toc_float: true
    code_folding: show
    code_download: true
    number-sections: true
---

# Setup

```{r setup, include=FALSE}
knitr::opts_chunk$set(
  echo = TRUE,
  warning = FALSE,
  message = FALSE,
  cache = TRUE
)
```

## Load Libraries

```{r load_libraries}
# Core tidyverse
library(tidyverse)
library(janitor)
library(lubridate)

# Spatial data
library(sf)
library(tigris)

# Census data
library(tidycensus)

# Weather data
library(riem)  # For Philadelphia weather from ASOS stations

# Visualization
library(viridis)
library(grid)
library(gridExtra)
library(knitr)
library(kableExtra)

library(ggplot2)
library(scales)
library(patchwork)

# here!
library(here)

# Extra
library(tidytext)
library(MASS)
library(jsonlite)
library(dplyr)
library(purrr)
library(spdep)
library(rsample)


# Get rid of scientific notation. We gotta look good!
options(scipen = 999)
```

## Set Census API Key

```{r census_key}
key <- Sys.getenv("CENSUS_API_KEY")
census_api_key(key)
```

# Load Spatial and System Data
```{r load_ss_data, message=FALSE, warning=FALSE, results='hide'}
# 1. Spatial Boundaries (Tracts)
tracts_sf <- tracts(state = "PA", county = "Philadelphia", year = 2021, cb = TRUE) |>
  st_transform(26918) |>
  dplyr::select(GEOID, geometry)

# 2. Bus Data by District
bus_district <- read_csv("data/Bus_Ridership_by_District.csv")

# 3. Transit Data by Mode
transit_mode <- read_csv("data/Avg_Daily_Ridership_By_Mode.csv")

# 4. Bus Shelters 
bus_shelters <- st_read("data/Bus_Shelters.geojson") |> st_transform(26918)

# 5. All Transit Stops
all_stops <- st_read("data/all_transit_stops.geojson") |>
  st_transform(26918)
```
::: callout-note
# Data Sources

- SEPTA Bus Ridership (APC): Average daily ridership by Census Tract/District (Source: OpenDataPhilly).

- SEPTA Bus Stops: Geospatial locations of all bus stops (Source: OpenDataPhilly).

- SEPTA Bus Shelters: Geospatial locations of all bus shelters (Source: OpenDataPhilly).

- SEPTA Transit Data by Mode: Average ridership for each mode (bus, trolley, rail, etc.) (Source: OpenDataPhilly).

- Census ACS (2017-2021): Demographic context (Poverty, Vehicle Ownership, Population Density) via tidycensus.

:::

# Load Ridership Data
```{r load_ridership_data}
# 1. Tract-Level Ridership (Raw)
bus_tract_raw <- read_csv("data/Bus_Ridership_by_Census_Tract.csv") |>
  clean_names() |>
  mutate(census_tract_id = as.character(census_tract_id))

# 2. Stop-Level Ridership (Raw List -> Combined Dataframe)
stops_files <- list.files("data", pattern = "Fall_.*\\.csv", full.names = TRUE)

# 3. Combine all bus stop files into a single df
bus_stops_raw <- stops_files |>
  map_dfr(function(file) {
    read_csv(file) |>
      mutate(year = as.integer(str_extract(file, "\\d{4}"))) |>
      clean_names() |>
      mutate(across(everything(), as.character)) 
  })
```

# Data Preprocessing and Cleaning

## Clean Ridership Data
```{r clean_ridership}
# A. Clean Tract Data
bus_tract_clean <- bus_tract_raw |>
  filter(county_name == "Philadelphia County") |>
  mutate(
    year = as.integer(str_extract(season, "\\d{4}")),
    total_ridership = on + off
  ) |>
  # Keep 2017+ for consistent analysis
  filter(year >= 2017)

# B. Clean Stop Data
bus_stops_clean <- bus_stops_raw |>
  type_convert() |> # Convert numbers back to numeric
  filter(year >= 2017) |>
  # Coalesce the mismatched column names from different years
  mutate(
    weekday_on  = coalesce(wkdy_ons, weekday_on),
    weekday_off = coalesce(wkdy_offs, weekday_of, weekday_of),
    saturday_on = coalesce(sat_ons, saturday_o),
    saturday_off = coalesce(sat_offs, saturday_1),
    sunday_on   = coalesce(sun_ons, sunday_ons),
    sunday_off  = coalesce(sun_offs, sunday_off),
    
    total_weekday = weekday_on + weekday_off
  ) |>
  # Keep only the clean columns
  dplyr::select(year, stop_code, stop_name = stop, route, total_weekday, 
         weekday_on, weekday_off, lat, lon)
```

## Filter Transit Stops (Bus vs. Rail/Trolley)
```{r clean_transit_stops}
# Define Rail/Trolley Codes to EXCLUDE
rail_codes <- c("BSL", "MFL", "HRS", "T1", "T2", "T3", "T4", "T5")

# 1. Rail Stops (for distance calc)
stops_rail <- all_stops |>
  filter(LineAbbr %in% rail_codes)

# 2. Bus Stops (Everything else, including T4 Bus, Owls, Trackless)
stops_bus <- all_stops |>
  filter(!LineAbbr %in% rail_codes)
```

# Feature Engineering

## Load Census Data
```{r load_census_data}
# ACS Demographics (2017-2021 5-Year)
acs_vars <- c(
  "B01001_001", # Total Population
  "B17001_002", # Poverty: Count below poverty level
  "B08201_002", # No Vehicle Available (Household)
  "B08201_001"  # Total Households
)

acs_data <- get_acs(
  geography = "tract", state = "PA", county = "Philadelphia",
  variables = acs_vars, year = 2021, output = "wide", geometry = FALSE
) |>
  dplyr::select(GEOID, 
         pop_total = B01001_001E,
         poverty_count = B17001_002E,
         no_vehicle_count = B08201_002E,
         hh_total = B08201_001E) |>
  mutate(
    pct_poverty = poverty_count / pop_total,
    pct_no_vehicle = no_vehicle_count / hh_total,
    across(c(pct_poverty, pct_no_vehicle), ~replace_na(., 0)) # Null handling
  )
```

## Spatial Engineering
```{r spatial_engineer}
# Count Shelters per Tract
shelters_agg <- tracts_sf |>
  st_join(bus_shelters) |>
  group_by(GEOID) |>
  summarise(n_shelters = n() - sum(is.na(objectid))) |> 
  st_drop_geometry()

# Count Bus Stops per Tract
stops_agg <- tracts_sf |>
  st_join(stops_bus) |>
  group_by(GEOID) |>
  summarise(n_stops = n()) |>
  st_drop_geometry()

# Distance to Nearest Rail Station
rail_union <- st_union(stops_rail)
tract_centroids <- st_centroid(tracts_sf)
dist_matrix <- st_distance(tract_centroids, rail_union)

dist_df <- tibble(
  GEOID = tracts_sf$GEOID,
  dist_to_rail_km = as.numeric(dist_matrix) / 1000
)
```

## Join Dataframes Together
```{r}
tract_features_static <- tracts_sf |>
  mutate(area_km2 = as.numeric(st_area(geometry)) / 1e6) |>
  st_drop_geometry() |>
  left_join(acs_data, by = "GEOID") |>
  left_join(shelters_agg, by = "GEOID") |>
  left_join(stops_agg, by = "GEOID") |>
  left_join(dist_df, by = "GEOID") |>
  mutate(
    n_stops = replace_na(n_stops, 0),
    n_shelters = replace_na(n_shelters, 0),
    stop_density = n_stops / area_km2,
    pop_density = pop_total / area_km2,
    # Equity Metric: Shelter Coverage
    pct_sheltered = ifelse(n_stops > 0, n_shelters / n_stops, 0),
    pct_sheltered = pmin(pct_sheltered, 1) # Cap at 100%
  )
```


# Exploratory Data Analysis

## Time Series Trends

### Average Daily Ridership by Mode

```{r mode_line_chart}
# Engineer a date feature for plotting
transit_mode_monthly <- transit_mode |>
  mutate(
    date = ymd(paste(Calendar_Year, Calendar_Month, "01", sep = "-"))
  )

# Plot line chart with a line for each transit mode over time
ggplot(transit_mode_monthly,
       aes(x = date,
           y = Average_Daily_Ridership,
           color = Mode)) +
  geom_line(linewidth = 1) +
  labs(
    title = "Average Daily Ridership by Transit Mode",
    subtitle = "SEPTA in Philadelphia from 2019 - 2025",
    x = "Date",
    y = "Average Daily Ridership",
    color = "Transit Mode"
  ) +
  theme_minimal()
```

### Ridership by Day of the Week

```{r ridership_dow}
# Aggregate to total ridership by year and day-of-week
bus_tract_year_dow <- bus_tract_clean |>
  group_by(year, day_of_week) |>
  summarise(
    total_ridership = sum(total_ridership, na.rm = TRUE),
    .groups = "drop"
  )

# Minimal line plot of total ridership over time by day-of-week
ggplot(bus_tract_year_dow,
       aes(x = year, y = total_ridership, color = day_of_week)) +
  geom_line(linewidth = 0.9) +
  geom_point(size = 1.8) +
  labs(
    title = "Total SEPTA Bus Ridership by Day of Week",
    subtitle = "Average daily boardings + departures for Fall only",
    x = "Year",
    y = "Total Average Daily Ridership",
    color = "Day of Week"
  ) +
  theme_minimal()

```

## Spatial Distributions

### Ridership in 2023

```{r}
# Prepare 2023 Tract Data from the CLEAN objects
ridership_2023 <- bus_tract_clean |>
  filter(day_of_week == "Weekday", year == 2023) |>
  group_by(census_tract_id) |>
  summarise(total_vol = sum(total_ridership, na.rm=TRUE))

# Join to Geometry
map_2023 <- tracts_sf |>
  left_join(ridership_2023, by = c("GEOID" = "census_tract_id")) |>
  mutate(
    # Create Quintiles
    vol_quintile = ntile(total_vol, 5),
    vol_quintile = factor(vol_quintile, labels = c("Lowest", "Low", "Med", "High", "Highest"))
  )

ggplot(map_2023) +
  geom_sf(aes(fill = vol_quintile), color = NA) +
  scale_fill_brewer(
    palette = "YlGnBu",
    na.value = "grey90",
    name = "Quintiles of Ridership"
  ) +
  labs(
    title = "Bus Ridership by Census Tract (Fall 2023)",
    subtitle = "Total Ridership = Boardings On_ + Off_",
  ) +
  theme_minimal() +
  theme(
  axis.text = element_blank(),
  axis.ticks = element_blank()
  )
```

## Equity Analysis
```{r poverty_vs_shelters}
# Join features to geometry for plotting
plot_data <- tracts_sf |>
  left_join(tract_features_static, by = "GEOID")

# Map A: Poverty
p1 <- ggplot(plot_data) +
  geom_sf(aes(fill = pct_poverty), color = NA) +
  scale_fill_viridis_c(option = "magma", labels = percent, name = "Poverty %") +
  labs(title = "Census Tract Poverty Rate (2021)", subtitle = "Where is the most financial distress?") +
  theme_void()

# Map B: Shelter Coverage
p2 <- ggplot(plot_data) +
  geom_sf(aes(fill = pct_sheltered), color = NA) +
  scale_fill_viridis_c(option = "viridis", labels = percent, name = "Sheltered %") +
  labs(title = "Percentage of Bus Stops with Shelters", subtitle = "Where are the highest % of shelters?") +
  theme_void()

# Display side-by-side
p1 + p2
```

```{r}
# Compute correlation between poverty and shelters
cor_val <- cor(tract_features_static$pct_poverty, tract_features_static$pct_sheltered, use="complete.obs")

# Plot scatterplot - No correlation
ggplot(tract_features_static, aes(x = pct_poverty, y = pct_sheltered)) +
  geom_point(alpha = 0.3) +
  geom_smooth(method = "lm", color = "red") +
  scale_x_continuous(labels = percent) +
  scale_y_continuous(labels = percent) +
  labs(title = paste0("Correlation(Poverty, Shelters): ", round(cor_val, 2)),
       subtitle = "Do poorer tracts have a smaller % of bus shelters?",
       x = "Poverty Rate", y = "Shelter Coverage")
```

##  Tract-Level Ridership Comparison (COVID Shock + Recovery)

```{r tract_perc_chg_19_21}
# 1. Prepare Data: Calculate Change for Both Periods
ridership_periods <- bus_tract_clean |>
  filter(
    year %in% c(2019, 2021, 2023),
    day_of_week == "Weekday"
  ) |>
  group_by(census_tract_id, year) |>
  summarise(total_ridership = sum(total_ridership, na.rm = TRUE), .groups = "drop") |>
  pivot_wider(
    names_from = year, 
    values_from = total_ridership, 
    names_prefix = "vol_"
  ) |>
  mutate(
    # Shock: 2019 -> 2021
    change_shock = (vol_2021 - vol_2019) / vol_2019,
    
    # Recovery: 2021 -> 2023
    change_recovery = (vol_2023 - vol_2021) / vol_2021,
    
    # Cap outliers for visualization (e.g., -100% to +100%)
    shock_capped = pmax(pmin(change_shock, 0.5), -1),
    recovery_capped = pmax(pmin(change_recovery, 1.0), -0.5)
  ) |>
  drop_na(change_shock, change_recovery)

# 2. Join to Geometry
map_periods <- tracts_sf |>
  left_join(ridership_periods, by = c("GEOID" = "census_tract_id"))

# 3. Create Plots

# Plot A: The COVID Shock (2019-2021)
p_shock <- ggplot(map_periods) +
  geom_sf(aes(fill = shock_capped), color = NA) +
  scale_fill_distiller(
    palette = "RdBu", direction = 1, 
    labels = scales::percent, 
    limits = c(-1, 0.5), # Fixed limits for comparison
    name = "Change"
  ) +
  labs(title = "The Shock (2019 → 2021)", 
       subtitle = "Ridership losses (pre-pandemic)") +
  theme_void() +
  theme(legend.position = "bottom")

# Plot B: The Recovery (2021-2023)
p_recovery <- ggplot(map_periods) +
  geom_sf(aes(fill = recovery_capped), color = NA) +
  scale_fill_distiller(
    palette = "RdBu", direction = 1, 
    labels = scales::percent,
    limits = c(-0.5, 1.0), # Different limits likely needed for recovery
    name = "Change"
  ) +
  labs(title = "The Recovery (2021 → 2023)", 
       subtitle = "Ridership gains (post-pandemic)") +
  theme_void() +
  theme(legend.position = "bottom")

# 4. Combine 
p_shock + p_recovery + 
  plot_annotation(
    title = "SEPTA Demand Resilience: Shock vs. Recovery",
    subtitle = "Comparing the spatial patterns of gains/losses due to the COVID pandemic",
    theme = theme(plot.title = element_text(size = 16, face = "bold"))
  )
```

## Fishnet Grid: Ridership + Bus Stops
```{r fishnet}
# 4. Fishnet Analysis (Stop Density & Weekday Ridership)

# 1. Prepare 2023 Data (Weekday Only)
stops_2023_sf <- bus_stops_clean |>
  filter(year == 2023) |>
  st_as_sf(coords = c("lon", "lat"), crs = 4326) |>
  st_transform(26918)

# 2. Create Fishnet Grid (500m)
# Use the union of tracts to define the city boundary
philly_boundary <- st_union(tracts_sf)

fishnet <- st_make_grid(
  philly_boundary,
  cellsize = 500, # 500m x 500m
  square = TRUE
) |>
  st_as_sf() |>
  mutate(grid_id = row_number())

# Clip grid to city shape (optional, makes map cleaner)
fishnet <- fishnet[philly_boundary, ]

# 3. Aggregate Stops to Grid
# Spatial join: Assign grid_id to each stop
stops_joined <- st_join(stops_2023_sf, fishnet, join = st_within)

# Summarize data by grid cell
fishnet_agg <- stops_joined |>
  st_drop_geometry() |> 
  group_by(grid_id) |>
  summarise(
    countStops = n(),
    totalRid = sum(total_weekday, na.rm = TRUE),
    .groups = "drop"
  )

# Join back to grid geometry for plotting
fishnet_plot_data <- fishnet |>
  left_join(fishnet_agg, by = "grid_id") |>
  # Replace NAs with 0 so empty cells plot as "0" instead of gray/missing
  mutate(
    countStops = replace_na(countStops, 0),
    totalRid   = replace_na(totalRid, 0)
  )

# 4. Plotting

# Plot A: Ridership Intensity
p_rid <- ggplot() +
  geom_sf(data = fishnet_plot_data, aes(fill = totalRid), color = NA) +
  geom_sf(data = philly_boundary, fill = NA, color = "white", size = 0.5) +
  scale_fill_viridis_c(
    name = "Weekday\nRidership",
    trans = "sqrt",
    option = "plasma"
  ) +
  labs(
    title = "Ridership Density",
    subtitle = "500m Grid, Fall 2023"
  ) +
  theme_void()

# Plot B: Stop Density
p_stops <- ggplot() +
  geom_sf(data = fishnet_plot_data, aes(fill = countStops), color = NA) +
  geom_sf(data = philly_boundary, fill = NA, color = "white", size = 0.5) +
  scale_fill_viridis_c(
    name = "Stop Count",
    trans = "sqrt",
    option = "viridis"
  ) +
  labs(
    title = "Stop Density",
    subtitle = "500m Grid, Fall 2023"
  ) +
  theme_void()

# Display side-by-side
p_rid + p_stops + 
  plot_annotation(
    title = "Comparison of Daily Weekday Ridership and Bus Stop Densities",
    theme = theme(plot.title = element_text(size = 16, face = "bold"))
  )
```

# Model Building Workflow

## Setup

```{r}
# 1. Prepare Longitudinal Panel (2017-2023)
# Aggregate ridership to one row per Tract-Year (Weekdays only)
bus_panel <- bus_tract_clean |>
  filter(day_of_week == "Weekday") |>
  group_by(census_tract_id, census_tract_name, year) |>
  summarise(total_ridership = sum(total_ridership, na.rm=TRUE), .groups="drop")

# 2. Join with Static Features (Demographics, Shelters, Rail Dist)
model_data <- bus_panel |>
  left_join(tract_features_static, by = c("census_tract_id" = "GEOID")) |>
  st_drop_geometry() |> # Drop geometry
  mutate(
    # Time Trends
    trend = year - min(year),
    post_covid = if_else(year >= 2021, 1, 0),
    
    # Log Transformations (handling zeros for density vars)
    log_pop_density = log(pop_density + 1),
    log_stop_density = log(stop_density + 1)
  ) |>
  filter(!is.na(total_ridership))
```


::: callout-note
# Key Modeling Decision: Only Predict Weekdays

- In our original modeling workflow, we attempted to model all day of week types found in the data: weekday, Saturday, and Sunday. However, we found that our model struggled to accurately predict for all three types in our initial findings.

- Even when incorporating dummy variables for each day of the week type, the model failed to learn differences in temporal patterns from those variables alone. In other words, it is not as simple to say that weekdays and weekends had the same travel patterns at different levels for the Fall season.

- Weekdays are driven by commuting for work, highlighting inelastic demand, where they must take the bus. Conversely, weekends are driven by discretionary travel where demand is highly elastic and irregular that isn't easily modelled from seasonal data alone. If we had access to daily data, it would be a much easier task to incorporate days of the week under the same model.

- As a result, weekdays have much more stable demand with less variation. However, for example, Sundays could be problematic to model with high variation due to weekend events such as Eagles games that we cannot directly capture from seasonal data. Even more broady, without daily data, a key factor like weather could not be used either.

- Thus, we came to the decision to only model for weekdays. Intuitively, this choice makes sense since the most regular transit patterns occur on weekdays with the most demand. Building a predictive model for weekdays is the better options to derive insights for meaningful policy considerations.
:::


```{r}
# 1. Define Error Metric
mae <- function(obs, pred) mean(abs(obs - pred))

# 2. Train/Test Split

# TRAIN = pre-COVID + two recovery year (2017-19, 2021-2022)
tract_train <- model_data |>
  filter(year != 2023) |>
  mutate(census_tract_id = as.factor(census_tract_id))

# TEST = heldout, evaluation year (2023)
tract_test <- model_data |>
  filter(year == 2023) |>
  mutate(census_tract_id = as.factor(census_tract_id))
```

## Baseline Model 1: Trend + COVID + Tract FE

```{r model_1}
# A. Fit Poisson
m1_pois <- glm(
  total_ridership ~ trend + post_covid + census_tract_id,
  data = tract_train,
  family = "poisson"
)

# B. Check Dispersion
dispersion <- sum(residuals(m1_pois, type = "pearson")^2) / m1_pois$df.residual

cat("Dispersion Parameter:", round(dispersion, 2), "\n")
if(dispersion > 1.5) cat("Overdispersion detected! Switching to Negative Binomial.\n")

# C. Fit Negative Binomial (The Real Baseline)
m1_nb <- glm.nb(
  total_ridership ~ trend + post_covid + census_tract_id,
  data = tract_train
)

# D. Evaluate
pred_m1 <- predict(m1_nb, newdata = tract_test, type = "response")
mae_m1  <- mae(tract_test$total_ridership, pred_m1)

cat("Baseline (M1) MAE:", round(mae_m1, 2))
```

## Model 2: Trend + COVID + Infrastructure + Equity

```{r model_2}
# Fit Model: Demographics + Infrastructure (No Tract FE)
m2_nb <- glm.nb(
  total_ridership ~ 
    trend + post_covid + 
    # Infrastructure Features
    stop_density + dist_to_rail_km + pct_sheltered +
    # Demographic Features
    pct_poverty + pct_no_vehicle + pop_density,
  data = tract_train
)

# Evaluate
pred_m2 <- predict(m2_nb, newdata = tract_test, type = "response")
mae_m2  <- mae(tract_test$total_ridership, pred_m2)

# Print Policy Findings
cat("M2 Equity + Infrastructure (No FE) MAE:", round(mae_m2, 2), "\n")
```

## Model 3: Trend + COVID + FE + COVID Interactions with Infrastructure/Equity
```{r model_3}
m3_nb <- glm.nb(
  total_ridership ~ 
    # 1. Global Trends
    trend + post_covid + 
    
    # 2. Differential Recovery (Interactions)
    # "Did high-poverty or sheltered areas recover differently?"
    post_covid:pct_poverty +
    post_covid:pct_sheltered +
    post_covid:dist_to_rail_km +
    
    # 3. Local Baselines (Fixed Effects)
    census_tract_id,
  data = tract_train
)

# Evaluate
pred_m3 <- predict(m3_nb, newdata = tract_test, type = "response")
mae_m3  <- mae(tract_test$total_ridership, pred_m3)

cat("M3 FE + Interactions MAE:", round(mae_m3, 2))
```
## Model 4: Equity Focused FE Model of Post-COVID Transit Recovery

::: callout-note
# Key Modeling Decision: Model 4

Model 4 uses a slightly smaller training dataset because it requires information from the previous year to construct lagged and interaction-based features (e.g., lag1_ridership and poverty_x_sheltered). Tracts without a prior-year observation cannot contribute meaningful values for these predictors, so they are excluded from training.

This is not a methodological inconsistency: all models are still evaluated on the same held-out 2023 test data, ensuring fair and comparable predictive accuracy. The reduced training sample reflects the structural requirements of the model, not selective filtering, and the improvement in accuracy demonstrates that incorporating prior-year trends and equity-focused interactions offers genuine predictive value.

:::

```{r add_features_m4}
model_data <- model_data %>%
  arrange(census_tract_id, year) %>%
  group_by(census_tract_id) %>%
  mutate(
    # Used only to ensure we have at least 1 prior year of data per tract
    lag1_ridership = lag(total_ridership, 1),
    
    # Equity–infrastructure interaction
    poverty_x_sheltered = pct_poverty * pct_sheltered
  ) %>%
  ungroup()
```

```{r model4_train_test}
tract_train_m4 <- model_data %>%
  filter(year != 2023) %>%
  filter(!is.na(lag1_ridership)) %>%          # <- key: only tracts with history
  mutate(census_tract_id = as.factor(census_tract_id))

tract_test_m4 <- model_data %>%
  filter(year == 2023) %>%
  filter(!is.na(lag1_ridership)) %>%
  mutate(census_tract_id = as.factor(census_tract_id))
```

```{r model4}
m4_nb <- glm.nb(
  total_ridership ~ 
    trend +
    post_covid +
    pct_poverty +
    pct_no_vehicle +
    pct_sheltered +
    poverty_x_sheltered +
    # key recovery test: do high-poverty tracts recover differently?
    post_covid:pct_poverty +
    census_tract_id,
  data = tract_train_m4
)

pred_m4 <- predict(m4_nb, newdata = tract_test_m4, type = "response")
mae_m4  <- mae(tract_test_m4$total_ridership, pred_m4)

cat("Model 4 (Best FE + Equity Recovery) MAE:", round(mae_m4, 2), "\n")

```


```{r model4_diagnostics_clean}

# Extract coefficient names
coef_all <- coef(m4_nb)

# Identify which are NOT tract fixed effects
# (FE all start with "census_tract_id")
coef_keep_idx <- !grepl("^census_tract_id", names(coef_all))

# Filtered coefficients
coef_main <- coef_all[coef_keep_idx]

cat("=== Model 4 Coefficients (Main Effects Only) ===\n")
print(round(coef_main, 4))
```


## Model 5: KNN 3 Bike stations 
```{r}
library(jsonlite)
library(dplyr)
library(purrr)
library(sf)

# 1. Pull Indego data from the API
url <- "https://kiosks.bicycletransit.workers.dev/phl"
indego_raw <- fromJSON(url, flatten = TRUE)

# This should be a data frame of all stations
indego_features <- indego_raw$features
# View the column names if you're curious:
# names(indego_features)

# 2. Extract lon/lat from geometry.coordinates (a list-column)
indego_df <- indego_features %>%
  mutate(
    lon = map_dbl(geometry.coordinates, ~ .x[1]),
    lat = map_dbl(geometry.coordinates, ~ .x[2])
  ) %>%
  transmute(
    kiosk_id    = properties.kioskId,
    name        = properties.name,
    total_docks = properties.totalDocks,
    lon, lat
  )

# 3. Convert to sf
indego_sf <- st_as_sf(indego_df, coords = c("lon", "lat"), crs = 4326)


```

```{r}
library(tigris)
options(tigris_use_cache = TRUE)

philly_tracts <- tracts("PA", "Philadelphia", year = 2020) %>%
  st_transform(4326)

indego_with_tracts <- st_join(indego_sf, philly_tracts["GEOID"])

```
```{r}
library(sf)
library(nngeo)

# Make sure both datasets are in projected CRS (meters)
tracts_sf <- st_transform(tracts_sf, 26918)
indego_sf <- st_transform(indego_sf, 26918)

# Tract centroids
tracts_centroids <- st_centroid(tracts_sf)

# KNN with k = 3
knn_res <- st_nn(
  tracts_centroids,
  indego_sf,
  k = 3,
  returnDist = TRUE
)

# Extract the distance matrix (rows = tracts, 3 columns = nearest stations)
dist_mat <- do.call(rbind, knn_res$dist)

# Create your ONE variable:
tracts_sf$knn3_mean_dist <- rowMeans(dist_mat)


```

```{r}
library(dplyr)

# 1. Build small table with tract ID + KNN value
tracts_knn <- tracts_sf |>
  st_drop_geometry() |>
  dplyr::select(GEOID, knn3_mean_dist)

# 2. Join onto model_data using correct ID mapping
model_data <- model_data |>
  dplyr::left_join(
    tracts_knn,
    by = c("census_tract_id" = "GEOID")
  )

# 3. Check results
head(model_data[, c("census_tract_id", "knn3_mean_dist")])


```

```{r}
# Rebuild train/test AFTER adding knn3_mean_dist to model_data
tract_train <- model_data |>
  filter(year != 2023) |>
  mutate(census_tract_id = as.factor(census_tract_id))

tract_test <- model_data |>
  filter(year == 2023) |>
  mutate(census_tract_id = as.factor(census_tract_id))

```


```{r}
library(MASS)

m5_nb <- glm.nb(
  total_ridership ~ 
    # 1. Global Trends
    trend + 
    post_covid + 
    
    # 2. Differential Recovery (Interactions)
    post_covid:pct_poverty +
    post_covid:pct_sheltered +
    post_covid:dist_to_rail_km +
    
    # NN3 Interaction
    post_covid:knn3_mean_dist +
    
    # 3. Local Baselines (Fixed Effects)
    census_tract_id,
  
  data = tract_train
)

# Evaluate
pred_m5 <- predict(m5_nb, newdata = tract_test, type = "response")
mae_m5  <- mae(tract_test$total_ridership, pred_m5)

cat("M5 FE + Interactions + KNN3 MAE:", round(mae_m5, 2), "\n")

```


## Model Performance Comparison

```{r model comparison}
# Helper metrics
mae <- function(obs, pred) mean(abs(obs - pred), na.rm = TRUE)

rmse <- function(obs, pred) sqrt(mean((obs - pred)^2, na.rm = TRUE))

mape <- function(obs, pred) {
  mask <- obs > 0  # avoid divide-by-zero issues
  mean(abs((obs[mask] - pred[mask]) / obs[mask]), na.rm = TRUE)
}

# Predictions on held-out 2023
pred_m1 <- predict(m1_nb, newdata = tract_test,   type = "response")
pred_m2 <- predict(m2_nb, newdata = tract_test,   type = "response")
pred_m3 <- predict(m3_nb, newdata = tract_test,   type = "response")
pred_m4 <- predict(m4_nb, newdata = tract_test_m4, type = "response")  # new Model 4
pred_m5 <- predict(m5_nb, newdata = tract_test,   type = "response")

# Observed outcomes
y_m1235 <- tract_test$total_ridership
y_m4_nb    <- tract_test_m4$total_ridership

# Comparison Table
results <- tibble(
  Model = c(
    "M1: Baseline (FE only)",
    "M2: Equity + Infrastructure (No FE)",
    "M3: FE + COVID Interactions",
    "M4: FE + Poverty Recovery (post_covid × poverty)",
    "M5: FE + COVID Interactions + KNN3"
  ),
  MAE = c(
    mae(y_m1235, pred_m1),
    mae(y_m1235, pred_m2),
    mae(y_m1235, pred_m3),
    mae(y_m4_nb,    pred_m4),
    mae(y_m1235, pred_m5)
  ),
  RMSE = c(
    rmse(y_m1235, pred_m1),
    rmse(y_m1235, pred_m2),
    rmse(y_m1235, pred_m3),
    rmse(y_m4_nb,    pred_m4),
    rmse(y_m1235, pred_m5)
  ),
  MAPE = c(
    mape(y_m1235, pred_m1),
    mape(y_m1235, pred_m2),
    mape(y_m1235, pred_m3),
    mape(y_m4_nb,    pred_m4),
    mape(y_m1235, pred_m5)
  ),
  AIC = c(
    AIC(m1_nb),
    AIC(m2_nb),
    AIC(m3_nb),
    AIC(m4_nb),      # updated to use new Model 4
    AIC(m5_nb)
  )
)

kable(results, digits = 3, caption = "Model Performance Comparison") |>
  kable_styling(bootstrap_options = c("striped", "condensed"), full_width = FALSE)
```

# Error Analysis

```{r m4_error_table}
# Recompute predictions to be safe
pred_m4 <- predict(m4_nb, newdata = tract_test_m4, type = "response")

errors_m4 <- tract_test_m4 %>%
  mutate(
    pred        = pred_m4,
    residual    = total_ridership - pred,             # actual - predicted
    abs_error   = abs(residual),
    pct_error   = if_else(total_ridership > 0,
                          residual / total_ridership, # signed %
                          NA_real_)
  )

# Quick numeric summary of error distribution
summary(errors_m4[, c("residual", "abs_error", "pct_error")])
```

### 5- Fold Cross Validation
```{r}
# Helper metrics
mae <- function(obs, pred) mean(abs(obs - pred), na.rm = TRUE)

rmse <- function(obs, pred) sqrt(mean((obs - pred)^2, na.rm = TRUE))

# Pre-2023 years only (your effective training window)
cv_years <- sort(unique(model_data$year[model_data$year != 2023]))

cv_years
```
```{r }
# Initialize tibble to store results
cv_results_m4_year <- tibble::tibble(
  holdout_year = integer(),
  MAE = numeric(),
  RMSE = numeric()
)
```



```{r}
# Loop over each holdout year
for (y in cv_years) {
  
  # 1. Training data:
  #    - years: all non-2023 years except the holdout year
  #    - require lag1_ridership (same restriction as Model 4)
  train_cv <- model_data %>%
    dplyr::filter(
      year != 2023,
      year != y,
      !is.na(lag1_ridership)
    ) %>%
    dplyr::mutate(
      census_tract_id = factor(census_tract_id)
    )
  
  # 2. Validation data:
  #    - year: the held-out year
  #    - require lag1_ridership
  valid_cv <- model_data %>%
    dplyr::filter(
      year == y,
      !is.na(lag1_ridership)
    ) %>%
    dplyr::mutate(
      # Force factor levels to match training data
      census_tract_id = factor(census_tract_id, levels = levels(train_cv$census_tract_id))
    ) %>%
    # Drop any tracts that never appear in training (would be NA after factoring)
    dplyr::filter(!is.na(census_tract_id))
  
  # 3. Fit Model 4 on training data
  m4_nb_cv <- MASS::glm.nb(
    total_ridership ~ 
      trend +
      post_covid +
      pct_poverty +
      pct_no_vehicle +
      pct_sheltered +
      poverty_x_sheltered +
      post_covid:pct_poverty +
      census_tract_id,
    data = train_cv
  )
  
  # 4. Predict onto held-out year
  preds <- predict(m4_nb_cv, newdata = valid_cv, type = "response")
  
  # 5. Compute metrics for this fold
  fold_mae  <- mae(valid_cv$total_ridership, preds)
  fold_rmse <- rmse(valid_cv$total_ridership, preds)
  
  # 6. Store results
  cv_results_m4_year <- dplyr::bind_rows(
    cv_results_m4_year,
    tibble::tibble(
      holdout_year = y,
      MAE  = fold_mae,
      RMSE = fold_rmse
    )
  )
}

cv_results_m4_year

```

```{r Mean CV Error}
# Overall average CV performance across years
cv_summary_m4_year <- cv_results_m4_year %>%
  dplyr::summarise(
    mean_MAE  = mean(MAE),
    sd_MAE    = sd(MAE),
    mean_RMSE = mean(RMSE),
    sd_RMSE   = sd(RMSE)
  )

cv_summary_m4_year
```




```{r m4_pred_vs_actual}
ggplot(errors_m4, aes(x = total_ridership, y = pred)) +
  geom_point(alpha = 0.5) +
  geom_abline(slope = 1, intercept = 0, linetype = "dashed") +
  scale_x_continuous(labels = scales::comma) +
  scale_y_continuous(labels = scales::comma) +
  labs(
    title = "Model 4: Predicted vs. Actual Ridership (2023 Test Year)",
    x = "Actual Total Ridership (Weekday, 2023)",
    y = "Predicted Total Ridership"
  ) +
  theme_minimal()
```

```{r m4_error_by_poverty}
errors_m4_equity <- errors_m4 %>%
  mutate(
    pov_quintile = ntile(pct_poverty, 5)
  ) %>%
  group_by(pov_quintile) %>%
  summarise(
    mean_poverty = mean(pct_poverty, na.rm = TRUE),
    MAE          = mean(abs_error, na.rm = TRUE),
    MAPE         = mean(abs(pct_error), na.rm = TRUE)
  )

kable(errors_m4_equity, digits = 3,
      col.names = c("Poverty Quintile", "Avg. Poverty Rate",
                    "MAE (Riders)", "MAPE")) %>%
  kable_styling(bootstrap_options = c("striped", "condensed"), full_width = FALSE)
```

```{r m4_error_by_poverty_plot}
ggplot(errors_m4_equity,
       aes(x = factor(pov_quintile), y = MAE)) +
  geom_col() +
  labs(
    title = "Model 4 Error by Poverty Quintile",
    x = "Poverty Quintile (1 = Lowest, 5 = Highest)",
    y = "Mean Absolute Error (Riders)"
  ) +
  theme_minimal()
```

```{r m4_errors_spatial_join}
errors_m4_sf <- tracts_sf %>%
  left_join(
    errors_m4 %>%
      dplyr::select(census_tract_id, residual, abs_error, pct_error),
    by = c("GEOID" = "census_tract_id")
  )
```

```{r m4_residual_map}
ggplot(errors_m4_sf) +
  geom_sf(aes(fill = residual), color = NA) +
  scale_fill_distiller(
    palette = "RdBu", direction = 1,
    name = "Residual (Actual - Predicted)"
  ) +
  labs(
    title = "Model 4 Residuals by Census Tract (2023)",
    subtitle = "Red = underpredicted (more riders than expected), Blue = overpredicted"
  ) +
  theme_void()
```

```{r m4_abs_error_map}
ggplot(errors_m4_sf) +
  geom_sf(aes(fill = abs_error), color = NA) +
  scale_fill_viridis_c(name = "Absolute Error") +
  labs(
    title = "Model 4 Absolute Error by Census Tract (2023)",
    subtitle = "Where is the model least accurate?"
  ) +
  theme_void()
```
```{r m4_morans_I}
# Keep only tracts with residuals
resid_data <- errors_m4_sf %>%
  filter(!is.na(residual))

coords <- st_coordinates(st_centroid(resid_data))

# k-nearest neighbors (k = 4)
nb <- knn2nb(knearneigh(coords, k = 4))
lw <- nb2listw(nb, style = "W")

moran_res <- moran.test(resid_data$residual, lw)

moran_res
```

::: callout-note
# Moran's I Interpretation

Moran’s I = 0.145 (z = 4.81, p < 0.001) indicates statistically significant positive spatial autocorrelation in the residuals. This means neighboring census tracts tend to have similar prediction errors, suggesting some spatial structure in ridership recovery that Model 4 does not fully capture.

:::
